---
title: "Taking Treatment Effects Analysis and Visualisation to Higher Levels with R"
output:
  prettydoc::html_pretty:
    theme: cayman #tactile, architect, leonids, hpstr
    highlight: vignette #github
    math: katex #latex
  html_document: default
  word_document: default
  pdf_document: 
    latex_engine: xelatex
description: |
  This blog is about a new function in my R package `Dyn4cast`. Again, there is a work horse behind the one line of code technology!
author:
  - name: Job Nmadu
    url: https://jobnmadu.com/
categories:
  - R
date: "2024-07-29T01:00:00Z"
tags: [RStats, Propensity Score, Treatment Effects, Data Science, Machine Learning]
---

# Introduction

This blog is a new function, `treatment_model` that have been added to the `Dyn4cast` package. The function provides means for enhanced estimation of propensity score and treatments effects from randomized controlled designed experiments.  

Observational study involves the evaluation of outcomes of participants not randomly assigned treatments or exposures. To be able to assess the effects of the outcome, the participants are matched using propensity scores (PSM). This then enables the determination of the effects of the treatments on those treated against those who were not treated. Most of the earlier functions available for this analysis only enables the determination of the average treatments effects on the treated (ATT) while the other treatment effects are optional. This is where this functions is unique because five different average treatment effects are estimated simultaneously, in spite of the _one line code_ arguments. The five treatment effects are:

Average treatment effect for the entire (ATE) population

Average treatment effect for the treated (ATT) population

Average treatment effect for the controlled (ATC) population

Average treatment effect for the evenly matched (ATM) population

Average treatment effect for the overlap (ATO) population.

There excellent materials dealing with each of the treatment effects, please [see](https://livefreeordichotomize.com/posts/2019-01-17-understanding-propensity-score-weighting/)

The basic usage of the codes are:

`treatment_model(Treatment, x_data)`  

Arguments

`Treatment` Vector of binary data (0, 1) LHS for the treatment effects estimation

`x_data` Data frame of explanatory variables for the RHS of the estimation

Let us go!

# Load library

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy      = "styler",
                      fig.align = "center",
                      comment   = NA,
                      warning   = FALSE,
                      error     = FALSE,
                      message   = FALSE,
                      collapse  = FALSE,
                      out.width = "100%",
                      dev = "ragg_png", #<- prevent default Windows device to
                      # render plots
                      dpi       = 132,
                      echo      = TRUE)
```


```{r}
library(Dyn4cast)
```

# Estimation of treatment effects model


```{r, warning=FALSE}

Treatment = treatments$treatment
data = treatments[, c(2:3)]
treat <- treatment_model(Treatment, data)

```

## Estimated treatment effects model

```{r, warning=FALSE}

summary(treat)
```

## Estimated various treatment effects

```{r, warning=FALSE}

treat$Effect
```

# Estimated propensity scores from the model

```{r}

head(treat$P_score)
```

##Ffitted values from the model

```{r}

head(treat$Fitted_estimate)
```

## Residuals of the estimated model

```{r}

tail(treat$Residuals)
```

## Plots of the propensity scores from the model

Quite a number of plots are in the works e.g. 

### Treatment effects (ATE)

```{r}

treat$`ATE plot`
```

### Treatment effects on the treated (ATT)

```{r}

treat$`ATT plot`
```

### Treatment effects on the evenly matched (ATM)

```{r}

treat$`ATM plot`
```

Welcome to the world of easy Data Science and easy Machine Learning!
