---
title: "Global movement of Happiness ladder with Machine learning in R"
output:
  prettydoc::html_pretty:
    theme: cayman #tactile, architect, leonids, hpstr
    highlight: vignette #github
    math: katex #latex
  html_document: default
  word_document: default
  pdf_document: 
    latex_engine: xelatex
description: |
  This blog is about the global happinness ladder capturing the changes in countries between 2002 and 2023.
author:
  - name: Job Nmadu
    url: https://jobnmadu.com/
categories:
  - R
date: "2024-07-26T01:00:00Z"
tags: [RStats, Data Science, Machine Learning, Factor analysis, Structural equation modelling]
---

# Introduction

This blog is about world happiness ladder using the world happiness report data sets (Helliwell et. al., 2024). The basic objective is to demonstrate the use of panel data which is quite distinct from cross-sectional or time series data. 

```{r setup, include=FALSE}

knitr::opts_chunk$set(tidy      = "styler",
                      fig.align = "center",
                      comment   = NA,
                      warning   = FALSE,
                      error     = FALSE,
                      message   = FALSE,
                      collapse  = FALSE,
                      out.width = "100%",
                      dev = "ragg_png", #<- prevent default Windows device to
                      # render plots
                      dpi       = 132,
                      echo      = FALSE)

options(tidyverse.quiet = TRUE)
library(psych)
library(tidyverse)
library(lavaan)
library(semPlot)
library(car)       # Companion to applied regression 
library(gplots)    # Various programing tools for plotting data
library(tseries)   # For timeseries analysis
library(lmtest)    # For hetoroskedasticity analysis
library(plm)
library(car)       # Companion to applied regression 
library(gplots)    # Various programing tools for plotting data
library(tseries)   # For timeseries analysis
library(lmtest)    # For hetoroskedasticity analysis
library(plm)

```


```{r}

Data <- read_csv("data/DataForTable2.1.csv")

Data$Regional <- case_match(Data$Region,
  "Australia and New Zealand"           ~ 1,
  "Central and Eastern Europe"          ~ 2,
  "Commonwealth of Independent States"  ~ 3,
  "East Asia"                           ~ 4,
  "Eastern Asia"                        ~ 4,
  "Latin America and Caribbean"         ~ 5,
  "Middle East and North Africa"        ~ 6,
  "Middle East and Northern Africa"     ~ 6,
  "North America"                       ~ 7,
  "North America and ANZ"               ~ 7,
  "South Asia"                          ~ 8,
  "Southern Asia"                       ~ 8,
  "Southeast Asia"                      ~ 9,
  "Southeastern Asia"                   ~ 9,
  "Sub-Saharan Africa"                  ~ 10,
  "Western Europe"                      ~ 11)

Data$Region1 <- case_match(Data$Region,
  "Australia and New Zealand"           ~ "Australiasia",
  "Central and Eastern Europe"          ~ "Central Europe",
  "Commonwealth of Independent States"  ~ "Com. Ind. States",
  "East Asia"                           ~ "East Asia",
  "Eastern Asia"                        ~ "East Asia",
  "Latin America and Caribbean"         ~ "Latin America",
  "Middle East and North Africa"        ~ "MENA",
  "Middle East and Northern Africa"     ~ "MENA",
  "North America"                       ~ "North America",
  "South Asia"                          ~ "South Asia",
  "Southern Asia"                       ~ "South Asia",
  "Southeast Asia"                      ~ "South East Asia",
  "Southeastern Asia"                   ~ "South East Asia",
  "Sub-Saharan Africa"                  ~ "SSA",
  "Western Europe"                      ~ "Western Europe")

ranked_data = Data %>%
  group_by(Year) %>%
  mutate(rank = min_rank(-Happiness), 
         Value_lbl = paste0(" ", Country, ": ",
                            format(Happiness, big.mark = ",",
                                   scientific = F)))%>%
  ungroup()

ranked_data_2023 <- ranked_data %>% 
  filter(Year == 2023)

year_data = Data %>%
  group_by(Year) %>%
  mutate(Mean = mean(Happiness)) %>%
  ungroup()

year_data_2023 <- year_data %>% 
  filter(Year == 2023)

tt_caption <- "Source: World Happiness Report  |  Graphics: Job Nmadu (@JobNmadu)"

```

# Global happiness ladder

## Cross-sectional happiness ladder for 2023

```{r}
ggplot(ranked_data_2023) +
  aes(x = Region1, y = Happiness, fill = Region1) +
  geom_col(position = "dodge") +
  theme_minimal() +
  coord_flip() +
  theme(legend.position = "none") +
 # facet_wrap(vars(Region1), scales = "free_y")+
  labs(x = "Region",
       y = "Happiness ladder",
       title = "Happiness order: 2023",
       subtitle = " ",
       #fill = "Regions",
       caption = tt_caption)
```

Fixed time, it is cross-sectional

## Times series vs. panel data visualisation

```{r}

ggplot(ranked_data) +
  aes(x = Year,
    y = Happiness,
    color = Region1) +
  #geom_area(size = 0.5) +
  geom_smooth(aes(color = Region1), method = "loess", se = FALSE,
              show.legend = TRUE) +
  scale_fill_hue(direction = 1) +
  scale_color_hue(direction = 1) +
  theme_minimal() +
 # theme(legend.position = "none") +
 # facet_wrap(vars(Region1), scales = "free_y")+
  labs(x = "Year",
       y = "Happiness ladder",
       title = "Happiness order: 2005 - 2023",
       subtitle = " ",
        color = "Regions",
       caption = tt_caption)

ggplot(ranked_data) +
  aes(x = Year,
    y = Happiness,
    fill = Region1,
    colour = Region1) +
  #geom_area(size = 0.5) +
  geom_smooth(aes(color = Region1), method = "loess", se = FALSE,
              show.legend = FALSE) +
  scale_fill_hue(direction = 1) +
  scale_color_hue(direction = 1) +
  theme_minimal() +
  theme(legend.position = "none") +
  facet_wrap(vars(Region1), scales = "free_y")+
  labs(x = "Year",
       y = "Happiness ladder",
       title = "Happiness order: 2005 - 2023",
       subtitle = " ",
       caption = tt_caption)
```

Each line is a timeseries but together, it is panel data

# Mean global happiness lalder

## Cross-sectional mean happiness ladder for 2023

```{r}
ggplot(year_data_2023) +
  aes(x = Region1, y = Happiness, fill = Region1) +
  geom_col(position = "dodge") +
  theme_minimal() +
  coord_flip() +
  theme(legend.position = "none") +
 # facet_wrap(vars(Region1), scales = "free_y")+
  labs(x = "Region",
       y = "Happiness ladder",
       title = "Mean happiness order: 2023",
       subtitle = " ",
      # fill = "Regions",
       caption = tt_caption)
```

Time is fixed, cross-sectional

## Times series vs. panel data visualisation

```{r}

ggplot(year_data) +
  aes(x = Year,
    y = Mean,
    color = Region1) +
  #geom_area(size = 0.5) +
  geom_smooth(aes(color = Region1), method = "loess", se = FALSE,
              show.legend = TRUE) +
  scale_fill_hue(direction = 1) +
  scale_color_hue(direction = 1) +
  theme_minimal() +
  #theme(legend.position = "none") +
  #facet_wrap(vars(Region1), scales = "free_y")+
  labs(x = "Year",
       y = "Happiness ladder",
       title = "Mean Happiness for individual regions: 2005 - 2023",
       subtitle = " ",
        color = "Regions",
       caption = tt_caption)

ggplot(year_data) +
  aes(x = Year,
    y = Mean,
    colour = Region1) +
  #geom_area(size = 0.5) +
  geom_smooth(aes(color = Region1), method = "loess", se = FALSE,
              show.legend = FALSE) +
  scale_fill_hue(direction = 1) +
  scale_color_hue(direction = 1) +
  theme_minimal() +
  #theme(legend.position = "none") +
  facet_wrap(vars(Region1), scales = "free_y")+
  labs(x = "Year",
       y = "Happiness ladder",
       title = "Mean Happiness for individual regions: 2005 - 2023",
       subtitle = " ",
       caption = tt_caption)
```

Each line is a timeseires but together, they are panel data

# Factor analysis of panel global happiness ladder

```{r}

source("docs/pca_summary.R")
source("docs/printloadings.R")

ll <- factoring(Data = Data[, -c(1:3,15)], exclude = 0)

printLoadings(ll$Extrated_factors$loadings)

fa.diagram(ll$Extrated_factors)

kableExtra::kable_styling(knitr::kable(ll$Parameters_of_Extracted_factos, digits = 2), "striped", position = "center", font_size = 12)

#load libraries
library(lavaan)
library(semPlot)
library(semTools)
library(nonnest2)
library(htmlTable)

#constructing the measurement model
mea_model <- 'MR1 =~ GDP + Life_Exp + Happiness + Support
              MR2 =~ Positive + Freedom + Generosity + Regional
              MR4 =~ Corruption 
              MR3 =~ Year '

#fitting the measurement model
mea_fit <- sem(mea_model,
               data = Data)

#summary statistics of the measurement model
summary(mea_fit,
        fit.measures = TRUE)

# 2.3 Obtain goodness of fit indicators of the model

kableExtra::kable_styling(knitr::kable(fitMeasures(mea_fit), rsqaure = TRUE), "striped", position = "center", font_size = 12)

kableExtra::kable_styling(knitr::kable(reliability(mea_fit)), "striped", position = "center", font_size = 12)

semPaths(mea_fit, "par", 
         edge.label.cex = 1.2, 
         fade = FALSE,
         edge.label.postition = 1,
         edge.label.font = 70,
         edge.label.by = TRUE, 
         layout = "tree",
         sizeMan = 10)

semPaths(mea_fit, what = "eq", whatLabels = "est", style = "lisrel",
         layout="spring", rotation=1, color = list(
           lat = rgb(255, 200, 135, maxColorValue = 255), 
           man = rgb(155, 240, 175, maxColorValue = 255)))
```

The inter-connectivity between the latent variables and the various variables used to measure hapiiness.


# References

Helliwell, J. F., Layard, R., Sachs, J. D., De Neve, J.-E., Aknin, L. B., & Wang, S. (Eds.). (2024). World Happiness Report 2024. University of Oxford: Wellbeing Research Centre.