---
title: "Revael the stories behind those Likert-type data"
output:
  prettydoc::html_pretty:
    theme: cayman #tactile, architect, leonids, hpstr
    highlight: vignette #github
    math: katex #latex
  html_document: default
  word_document: default
  pdf_document: 
    latex_engine: xelatex
description: |
  This blog is about two new functions in the R package `Dyn4cast` . The power of one line of code technology is still maintained..
author:
  - name: Job Nmadu
    url: https://jobnmadu.github.io/Dyn4cast/
categories:
  - R
date: "2023-09-29T01:00:00Z"
tags: [RStats, Garrett Ranking, Latent factors, Data Science, Machine Learning]
---

# Introduction

This blog is about two new functions, `Model_factors` and `garrett_ranking` that have been added to the `Dyn4cast` package. The two functions provides means for gaining deeper insights into the meaning behind Likert-type variables collected from respondents. Garrett ranking provides the ranks of the observations of the variables based on the level of seriousness attached to it by the respondents. On the other hand, Model factors determines and retrieve the latent factors inherent in such data which now becomes continuous data. The factors or data frame retrieved from the variables can be used in other analysis like regression and machine learning. 

The two functions are path of factor analysis, essentially, exploratory factor analysis (EFA), used to unravel the underlying structure of the observed variables. The analysis also helps to reduce the complex structure by determining a smaller number of latent factors that sufficiently represent the variation in the observed variables. With EFA, no prior knowledge or hypothesis about the number or nature of the factors is assumed. These are great tools to help tell the story behind your data. The data used for `Model_factors` is prepared using `fa.parallel` and `fa` functions in the `psych` package. The interesting thing about these functions are their simplicity, and we still maintain the _one line_ code technique.

The basic usage of the codes are:

`garrett_ranking(data, num_rank, ranking = NULL, m_rank = c(2:15))`

`Data` The data for the Garrett Ranking.

`num_rank` number of ranks applied to the data. If the data is a five-point Likert-type data, then number of ranks is 5.

`Ranking` A vector of list representing the ranks applied to the data. If not available, positional ranks are applied.

`m_rank` scope of ranking (2-15).

`Model_factors(data = dat, DATA = Data)`

`data` `R` object⁠ obtained from EFA using the `fa` function in `psych` package

`DATA` `data.frame` of the raw data used to obtain data object.


Let us go!

# Load library

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy      = "styler",
                      fig.align = "center",
                      comment   = NA,
                      warning   = FALSE,
                      error     = FALSE,
                      message   = FALSE,
                      collapse  = FALSE,
                      out.width = "100%",
                      dev = "ragg_png", #<- prevent default Windows device to
                      # render plots
                      dpi       = 132,
                      echo      = TRUE)
```


```{r}
library(Dyn4cast)
```

# Garrett Ranking

## ranking is supplied

```{r, warning=FALSE}

garrett_data <- data.frame(garrett_data)
ranking <- c("Serious constraint", "Constraint",
"Not certain it is a constraint", "Not a constraint",
"Not a serious constraint")

garrett_ranking(garrett_data, 5, ranking)
```

## ranking not supplied

```{r, warning=FALSE}
garrett_ranking(garrett_data, 5)
```

# you can rank subset of the data

```{r, warning=FALSE}

garrett_ranking(garrett_data, 8)

garrett_ranking(garrett_data, 4)
```

# Latent Variables Recovery

```{r}
library(psych)
Data <- Quicksummary
GGn <- names(Data)
GG <- ncol(Data)
GGx <- c(paste0('x0', 1:9), paste("x", 10:ncol(Data), sep = ""))
names(Data) <- GGx
lll <- fa.parallel(Data, fm = 'minres', fa = 'fa')
dat <- fa(Data, nfactors = lll[["nfact"]], rotate = "varimax",fm="minres")

DD <- Model_factors(data = dat, DATA = Data)

DD$Latent_1

DD$Latent_3

DD$Latent_5

```


Welcome to the world of Data Science and easy Machine Learning!
